---
title: "SCT Task 2"
author: "Michael Bohl"
date: "2024-04-24"
output: html_document
---

## IMPORTS, install libraries if needed
```{r}
required_libraries = c("BiocManager", "R.utils", "Seurat", "Matrix", "dplyr", "patchwork", "devtools", "simspec", "ggplot2", "ggrepel", "DESeq2", "scran", "pbapply", 'patchwork') # If something doesnt work, install through BiocManager::install("package_name")

# load libraries and install from CRAN if necessary
for (library_name in required_libraries) {
  if (!requireNamespace(library_name, quietly = TRUE)) {
    message(paste("Installing and loading library", library_name))
    install.packages(library_name)
  }
  library(library_name, character.only = TRUE)
}

# libraries that are not available from CRAN
# BiocManager::install("Seurat")
# BiocManager::install("DESeq2")
# BiocManager::install("scran")
# devtools::install_github('quadbio/simspec')
#remotes::install_github('chris-mcginnis-ucsf/DoubletFinder') # To filter out doublets in the data
```

### Part 1. Descriptive analysis of the data set


## read in the data (rename the counts file to matrix.mtx.gtz)
```{r}
counts <- Read10X("./data", gene.column=1)
meta <- read.table("./data/meta.tsv.gz")
seurat <- CreateSeuratObject(counts, meta.data = meta)
```

## Quality control

  # Cells with too few genes detected
  # Cells with too many genes detected
  # Cells with high mitochondrial transcript percentage

```{r}
seurat[["percent.mt"]] <- PercentageFeatureSet(seurat, pattern = "^MT[-\\.]")
# Combined for all organoids
VlnPlot(seurat, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
VlnPlot(seurat, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0)
# Separate
VlnPlot(seurat, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, group.by = 'organoid')
VlnPlot(seurat, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, group.by = 'organoid')
```
  # Gene number and transcript correlation
```{r}
plot1 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "percent.mt", group.by = 'organoid')
plot2 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", group.by = 'organoid')
plot1 + plot2 # Very high correlation of transcript nr and gene nr.
# Due to the correlation of gene number and transcript number, we only need to set a cutoff to either one of these metrics, combined with an upper threshold of mitochondrial transcript percentage, for the QC.
```
  # Subset
```{r}
seurat <- subset(seurat, subset = nFeature_RNA > 500 & nFeature_RNA < 4500 & percent.mt < 5.5)
```

## routine processing
```{r}
# Normalize, find variable features, remove cell cycle genes, and scale data
seurat <- NormalizeData(seurat) %>%
  FindVariableFeatures(nfeatures = 3000) %>%
  CellCycleScoring(s.features = cc.genes.updated.2019$s.genes, g2m.features = cc.genes.updated.2019$g2m.genes)
VariableFeatures(seurat) <- setdiff(VariableFeatures(seurat), c(unlist(cc.genes.updated.2019), grep("^MT-", rownames(seurat), value=T))) # remove cell cycle genes and mitochondrial genes
```

  # Elbow plot to determine PCA number
```{r}
#seurat <- RunPCA(seurat, npcs = 50)
#ElbowPlot(seurat, ndims = ncol(Embeddings(seurat, "pca"))) # To be fair, 16 looks better as the latter 4 look to be on a plateau
```


```{r}
seurat <- ScaleData(seurat, vars.to.regress = c("G2M.Score","S.Score")) %>% # regress out cell cycle scores
  RunPCA(npcs = 20) %>%
  RunUMAP(dims = 1:20) 

plot1 <- UMAPPlot(seurat, group.by="organoid") & NoAxes()
plot2 <- FeaturePlot(seurat, c("FOXG1","EMX1","DLX2","MKI67"), order=T) & NoAxes() & NoLegend()
plot1 | plot2
```  

  # Filter out Doublets
```{r}
# This requires an identified pK, because of time constraints I leave this empty for now. ~ Aurimas
# If anyone wants to, refer to ?DoubletFinder
```

  # Final visualisation before integration
```{r}
seurat <- RunTSNE(seurat, dims = 1:20)
seurat <- RunUMAP(seurat, dims = 1:20)

plot1 <- TSNEPlot(seurat, group.by = 'organoid')
plot2 <- UMAPPlot(seurat, group.by = 'organoid')

plot1 
plot2 # Some integration would be helpful looks like
```

# Store seurat for integration
```{r}
seurat_raw <- seurat
```



## CSS integration
```{r}
seurat <- cluster_sim_spectrum(seurat, label_tag = "organoid", cluster_resolution = 1) 
seurat <- RunUMAP(seurat, reduction = "css", dims = 1:ncol(Embeddings(seurat,"css")), reduction.name = "umap_css", reduction.key = "UMAPCSS_")
plot1 <- DimPlot(seurat, reduction = "umap_css", group.by="GLI3_status") & NoAxes()
plot1 <- DimPlot(seurat, reduction = "umap_css", group.by="organoid") & NoAxes()
plot2 <- FeaturePlot(seurat, c("FOXG1","EMX1","DLX2","MKI67"), reduction = "umap_css", order=T) & NoAxes() & NoLegend()
plot1 | plot2
```


## Different integration methods
  # TODO: Fix: CCAIntegration, RPCAIntegration, HarmonyIntegration, FastMNNIntegration, scVIIntegration
```{r}
# Define integration methods
integration_methods <- c("CCAIntegration", "RPCAIntegration", "HarmonyIntegration", 
                          "FastMNNIntegration", "scVIIntegration")

integrated_objects <- list()

for (method in integration_methods) {
  integrated_object <- IntegrateLayers(
    object = seurat_raw,
    method = method,
    orig.reduction = "pca",
    
    verbose = FALSE
  )
  integrated_objects[[method]] <- integrated_object
}
```
  # Visualize integration methods
```{r}
integration_plots <- list()
for (seurat_obj in integrated_objects) {
  seurat_obj <- FindNeighbors(seurat_obj, reduction = "integrated.cca", dims = 1:20)
  seurat_obj <- FindClusters(obj, resolution = 1.2, cluster.name = "cca_clusters") # Random resolution
  # Store UMAP plots for each of the seurat objects into a list
  
}

for (){
  seurat_obj <- RunUMAP(seurat_obj, reduction = "integrated.cca", dims = 1:20)
}
```
  # Comprehensive feature plot
```{r}
# UNCOMMENT ONLY IF YOU WANT TO SAVE THE PDF
# Create a PDF file to save the plots
pdf("gene_plots.pdf")

cell_markers <- c(
  "POU5F1",  # pluripotency
  "PAX6", "VIM",  # neural progenitor cell
  "EMX1", "NEUROD6",  # dorsal telencephalon
  "DLX5", "ISL1", "GAD1",  # ventral telencephalon
  "TCF7L2", "LHX9",  # non-telencephalic regions
  "DCN", "COL5A1",  # mesenchymal population
  "SIX3", "CDH2", "SOX3", "HES5",  # neuroectodermal population
  "DLX5", "TFAP2A",  # non-neural ectoderm markers
  "LDHA",  # neuroepithelial population
  "FGF8", "BMP7", "EMX1", "DLX2",  # telencephalic NPCs
  "WLS", "WNT8B",  # non-telencephalic NPCs
  "SIX6", "HES5", "SIX3", "FGFR3", "RSPO3", "WNT7B", "MSX1", "BAMBI", "BNC2"  # patterning center markers and self-patterning variation markers
)

# Loop through each gene
for (gene in cell_markers) {
  
  # Create FeaturePlot for tSNE reduction
  #plot_tsne <- FeaturePlot(seurat, gene, reduction = "tsne") # I use either or, so can choose tsne if you prefer. ~ A

  # Create FeaturePlot for UMAP reduction
  plot_umap <- FeaturePlot(seurat, gene, reduction = "umap_css")

  # Save the plots on separate pages
  #plot(plot_tsne)
  plot(plot_umap)
}

plot1 <- DimPlot(seurat, reduction = "umap_css", label = T) & NoAxes()
plot(plot1)
# Close the PDF device
dev.off()
```

  # Manual-clustering code
```{r}
# Duplicate Seurat obj. to not lose it#
#seurat_obj <- seurat


```


```{r}
# TODO: annotate clusters, this is too inaccurate
seurat@meta.data$region <- factor(setNames(c("ventral telen.",
                                   "ventral telen.",
                                   "non-telen.",
                                   "dorsal telen.",
                                   "non-telen.",
                                   "non-telen.",
                                   "non-telen.",
                                   "ventral telen.",
                                   "non-telen.",
                                   "dorsal telen.",
                                   "non-telen.",
                                   "ventral telen.",
                                   "non-telen.",
                                   "non-telen.",
                                   "non-telen.",
                                   "mixed",
                                   "dorsal telen.",
                                   "non-telen.",
                                   "dorsal telen.",
                                   "non-telen.",
                                   "dorsal telen."),
                                 levels(seurat@active.ident))[seurat@active.ident],
                        levels=c("dorsal telen.","ventral telen.","non-telen.","mixed"))
plot1 <- DimPlot(seurat, reduction = "umap_css", group.by = "region") & NoAxes()
plot2 <- FeaturePlot(seurat, c("SOX2","DCX","FOXG1","EMX1","DLX2","MKI67","OTX2","HOXB2","TTR"), reduction = "umap_css", order=T) & NoAxes() & NoLegend()
plot1 | plot2
```

### Part 2. Comparison of cell type compositions between conditions

## visually check cell type compositions
```{r}
freq <- table(seurat$region, seurat$organoid) # create a frequency table of annotated cell types
prop <- apply(freq,2,function(x) x/sum(x)) # calculate the proportion of each cell type

layout(matrix(1:3,nrow=1)); par(mar=c(8,5,1,1)) # set the layout for the plots
barplot(freq, col=c("#E74C3C","#A569BD","#76D7C4","#CCD1D1"),
        border=NA, las=2, ylab="Frequency", cex.names = 0.8)
barplot(prop, col=c("#E74C3C","#A569BD","#76D7C4","#CCD1D1"),
        border=NA, las=2, ylab="Proportion", cex.names = 0.8)
plot.new()
legend("left", fill=c("#E74C3C","#A569BD","#76D7C4","#CCD1D1"), legend=rownames(freq), bty="n")

```
# visible difference between WT and KO: ... (fix once clusters are annotated)


## compare cells using Fisher's exact test
```{r}
freq_fisher <- function(conditions, have_identity){
  freq <- table(factor(have_identity, levels=c(TRUE,FALSE)),
                conditions)
  test <- fisher.test(freq)
  res <- setNames(c(test$estimate, test$p.value), c("oddsratio","pval_fisher"))
  return(res)
}

region_enrichment <- data.frame(region = levels(seurat$region),
                                # loop and test every region
                                t(sapply(levels(seurat$region), function(region)
                                  freq_fisher(conditions = factor(seurat$GLI3_status, levels=c("KO","WT")),
                                              have_identity = seurat$region == region)
                                )),
                                row.names=NULL)
region_enrichment$padj_fisher <- p.adjust(region_enrichment$pval_fisher)

region_enrichment
```
