---
title: "SCT Task 2"
author: "Michael Bohl"
date: "2024-04-24"
output: html_document
---

## IMPORTS, install libraries if needed
```{r}
required_libraries = c("BiocManager", "R.utils", "Seurat", "Matrix", "dplyr", "patchwork", "devtools", "simspec", "ggplot2", "ggrepel", "DESeq2", "scran", "pbapply")

# load libraries and install from CRAN if necessary
for (library_name in required_libraries) {
  if (!requireNamespace(library_name, quietly = TRUE)) {
    message(paste("Installing and loading library", library_name))
    install.packages(library_name)
  }
  library(library_name, character.only = TRUE)
}

# libraries that are not available from CRAN
# BiocManager::install("Seurat")
# BiocManager::install("DESeq2")
# BiocManager::install("scran")
# devtools::install_github('quadbio/simspec')

```

### Part 1. Descriptive analysis of the data set


## read in the data (rename the counts file to matrix.mtx.gtz)
```{r}
counts <- Read10X("./data", gene.column=1)
meta <- read.table("./data/meta.tsv.gz")
seurat <- CreateSeuratObject(counts, meta.data = meta)
```


## routine processing
```{r}
# Normalize, find variable features, remove cell cycle genes, and scale data
seurat <- NormalizeData(seurat) %>%
  FindVariableFeatures(nfeatures = 3000) %>%
  CellCycleScoring(s.features = cc.genes.updated.2019$s.genes, g2m.features = cc.genes.updated.2019$g2m.genes)
VariableFeatures(seurat) <- setdiff(VariableFeatures(seurat), c(unlist(cc.genes.updated.2019), grep("^MT-", rownames(seurat), value=T))) # remove cell cycle genes and mitochondrial genes

seurat <- ScaleData(seurat, vars.to.regress = c("G2M.Score","S.Score")) %>% # regress out cell cycle scores
  RunPCA(npcs = 20) %>%
  RunUMAP(dims = 1:20) 

plot1 <- UMAPPlot(seurat, group.by="organoid") & NoAxes()
plot2 <- FeaturePlot(seurat, c("FOXG1","EMX1","DLX2","MKI67"), order=T) & NoAxes() & NoLegend()
plot1 | plot2
```  


## CSS integration
```{r}
seurat <- cluster_sim_spectrum(seurat, label_tag = "organoid", cluster_resolution = 1) 
seurat <- RunUMAP(seurat, reduction = "css", dims = 1:ncol(Embeddings(seurat,"css")), reduction.name = "umap_css", reduction.key = "UMAPCSS_")
plot1 <- DimPlot(seurat, reduction = "umap_css", group.by="GLI3_status") & NoAxes()
plot1 <- DimPlot(seurat, reduction = "umap_css", group.by="organoid") & NoAxes()
plot2 <- FeaturePlot(seurat, c("FOXG1","EMX1","DLX2","MKI67"), reduction = "umap_css", order=T) & NoAxes() & NoLegend()
plot1 | plot2
```


## Cluster and annotate data
```{r}
seurat <- FindNeighbors(seurat, reduction = "css", dims = 1:ncol(Embeddings(seurat,"css"))) %>%
  FindClusters(resolution = 0.9)

plot1 <- DimPlot(seurat, reduction = "umap_css", label = T) & NoAxes()
plot2 <- FeaturePlot(seurat, c("SOX2","DCX","FOXG1","EMX1","DLX2","MKI67","OTX2","HOXB2","TTR"),
                     reduction = "umap_css", order=T) & NoAxes() & NoLegend()
plot1 | plot2

# TODO: annotate clusters, this is too inaccurate
seurat@meta.data$region <- factor(setNames(c("ventral telen.",
                                   "ventral telen.",
                                   "non-telen.",
                                   "dorsal telen.",
                                   "non-telen.",
                                   "non-telen.",
                                   "non-telen.",
                                   "ventral telen.",
                                   "non-telen.",
                                   "dorsal telen.",
                                   "non-telen.",
                                   "ventral telen.",
                                   "non-telen.",
                                   "non-telen.",
                                   "non-telen.",
                                   "mixed",
                                   "dorsal telen.",
                                   "non-telen.",
                                   "dorsal telen.",
                                   "non-telen.",
                                   "dorsal telen."),
                                 levels(seurat@active.ident))[seurat@active.ident],
                        levels=c("dorsal telen.","ventral telen.","non-telen.","mixed"))
plot1 <- DimPlot(seurat, reduction = "umap_css", group.by = "region") & NoAxes()
plot2 <- FeaturePlot(seurat, c("SOX2","DCX","FOXG1","EMX1","DLX2","MKI67","OTX2","HOXB2","TTR"), reduction = "umap_css", order=T) & NoAxes() & NoLegend()
plot1 | plot2
```

### Part 2. Comparison of cell type compositions between conditions

## visually check cell type compositions
```{r}
freq <- table(seurat$region, seurat$organoid) # create a frequency table of annotated cell types
prop <- apply(freq,2,function(x) x/sum(x)) # calculate the proportion of each cell type

layout(matrix(1:3,nrow=1)); par(mar=c(8,5,1,1)) # set the layout for the plots
barplot(freq, col=c("#E74C3C","#A569BD","#76D7C4","#CCD1D1"),
        border=NA, las=2, ylab="Frequency", cex.names = 0.8)
barplot(prop, col=c("#E74C3C","#A569BD","#76D7C4","#CCD1D1"),
        border=NA, las=2, ylab="Proportion", cex.names = 0.8)
plot.new()
legend("left", fill=c("#E74C3C","#A569BD","#76D7C4","#CCD1D1"), legend=rownames(freq), bty="n")

```
# visible difference between WT and KO: ... (fix once clusters are annotated)


## compare cells using Fisher's exact test
```{r}
freq_fisher <- function(conditions, have_identity){
  freq <- table(factor(have_identity, levels=c(TRUE,FALSE)),
                conditions)
  test <- fisher.test(freq)
  res <- setNames(c(test$estimate, test$p.value), c("oddsratio","pval_fisher"))
  return(res)
}

region_enrichment <- data.frame(region = levels(seurat$region),
                                # loop and test every region
                                t(sapply(levels(seurat$region), function(region)
                                  freq_fisher(conditions = factor(seurat$GLI3_status, levels=c("KO","WT")),
                                              have_identity = seurat$region == region)
                                )),
                                row.names=NULL)
region_enrichment$padj_fisher <- p.adjust(region_enrichment$pval_fisher)

region_enrichment
```
